Accessible, realistic genome simulation with selection using \stdpopsim

### 2025-07-08 review-responses.tex

  \begin{point}{\revref} % line 577
    Sentence starting on line 383 about difficulty identifying h and s separately seems to be missing some important citations, 
    e.g. Fuller et al 2019 (pmid 30962618) and citations therein, including Simons et al 2014; Wright 1937.
  \end{point}

  \reply{
    TODO. \revref
  }

  \begin{point}{} % line 577
    It might bear further comment that the gamma distribution is not flexible enough to accommodate the distribution of hs in the vaquita simulation 
    (which is what the DFE inference methods can really see), even if the s values are gamma distributed (bc of the big jump caused in hs by the switch in dominance). 
    This is interesting given how often gammas are used in DFE inference. The paper does mention that using h=1/2 as an assumption can lead to bad estimation of s, 
    but the point that a relationship between s and h can cause hs to be non-gamma even when s is gamma is perhaps a little more subtle. 
    Generally, I thought the lessons learned from the DFE exercise here could be put more clearly.
  \end{point}

  \reply{
    Following this insightful comment, we added a few sentences to the discussion addressing this issue \revref:
    %IG: I added this quote of the text I added to the discussion to clarify our response to the reviewer.
    %    We'll need to update after we finalize the version in the revision.
    \textit{``
    Our analysis suggests that the distributions of $s$ estimated by these methods
    provide a reasonable approximation for the true distribution of $hs$ (Figure \ref{fig:vaquita-dfe}C).
    This points to a notable limitation of most existing DFE inference methods,
    which assume that selection coefficients are gamma-distributed.
    % DRS: see comment in main text on whether the limitation is the gamma for s or the fact that h is fixed
    As shown in Figure \ref{fig:vaquita-dfe}C,
    while a gamma distribution might fit the distribution of $s$, a relationship between $h$ and $s$ might result in
    $hs$ which is not well approximated by a gamma distribution.
    "}
  }

@@

  \begin{point}{}
    The inference of the deleterious DFE with GRAPES is quite inaccurate. 
    Is it possible that there is a factor of 2 error involved?
  \end{point}

  \reply{
As noted in the paper, conversion of selection inferences between tools is challenging.
We thus consulted with the developers of \polydfe and \grapes to confirm our conversions.
(The developer of \dadicli is an author on this paper.)
The inaccuracy of \grapes is a surprise to us.
Note that \grapes is also less accurate than \polydfe and \dadicli in inferring the shape of the distribution (Fig.~\ref{fig:homsap-dfe.ooa}B), which suggests the inaccuracy is not simply a scaling issue.
  }

  \begin{point}{\revref} % 21:
    Figure 3c does not have any error bars, it’s difficult to conclude if the inference is actually significantly different from the truth. 
    Which “N” was used to obtain estimates of “s” in Figure 3a? 
    This should be mentioned in the legend.  \end{point}

  \reply{
  %IG: I don't think that error bars are appropriate here and I tried to explain why
  The main purpose of the plots in Figure 3C is to show the actual differences between simulated
  and inferred distributions of $s$, since differences in mean and shape (shown in panels A-B) are not
  easy to judge assess what practical impact the parameter differences have on the realized distributions. 
  We have added standard error bars to the mean values in Figure 3C, but the variation among the two parameters themselves 
  is better for directly assessing the differences among methods. 
  Regarding the estimates of $\text{E}[s]$ shown in Figure 3A (and also Figure 4A and associated supplementary figures),
  these values were produced directly by the inference methods, which output their own estimate of $\theta = 4N_{e}\mu$,
  where $\mu$ is taken from as the value in the stdpopsim catalog for each species. 
  We have clarified this in the main text.
  In the last paragraph of the methods section on DFE inference, we provide a complete description of the scaling
  we apply to the output of each method so that they are directly comparable to $s$ as simulated in \slim.
  }

### 2025-07-08 analysis2_manuscript.tex
fig3C error bars and runtime fig.

    % sweep interface
    While the \texttt{DFE} class facilitates the simulation of non-neutral variation across a sequence,
    it does not provide a means of specifying the origin and fate of particular selected alleles.
    Thus, we have also implemented an API in \stdpopsim (\texttt{stdpopsim.selective\_sweep}) that enables 
    selective sweeps to be simulated with a high degree of control.
    This function augments a demographic model by introducing an advantageous mutation over a given time interval,
    at a given position along the sequence; and conditions the simulation such that the mutation is above a particular 
    frequency at the beginning and end of the sweep period.
    The conditioning is implemented via rejection sampling; if the frequency of the advantageous allele 
    drifts out of the prescribed bounds, the simulation is restarted from the introduction of 
    the advantageous allele. If there are competing advantageous alleles (e.g. simultaneous sweeps at distinct loci) 
    then the simulation is restarted from the introduction of the first of these.
    As this functionality is implemented on top of the existing \stdpopsim API, it may be combined with other 
    models of selection and demography to simulate the combined effects of multiple, disparate evolutionary processes
    on genetic variation in a population.

@@

    \begin{figure}[b!]
        \centering
        \includegraphics[width=\linewidth]{figures/HomSap/OOA/HomSap_discrete_DFE}
        \caption{Inferred versus simulated DFEs from simulated human genomes.
        Simulations were performed using a human out-of-Africa demographic model with a gamma-distributed DFE
        acting on exons (see \textbf{Methods}).
        DFE is inferred separately for each of the three extant populations (CEU, CHB, and YRI)
        by three different methods: \grapes, \polydfe , and \dadicli.
        (A-B) Mean absolute value of selection coefficient ($\lvert E(s) \rvert $) and shape parameter are
        shown for each DFE inferred from all three simulated datasets,
        with median values marked by horizontal bars
        and simulated values represented by dashed horizontal lines.
        (C) Binned distribution of $s$ implied by the average DFE inferred from the three simulated datasets (averaging the inferred gamma parameters);
        white bars represent the distribution used in the simulation. 
        }
        \label{fig:homsap-dfe.ooa}
    \end{figure}

@@

    \begin{figure}[b!]
        \centering
        \includegraphics[width=0.8\textwidth]{figures/PhoSin/Vaquita2Epoch_1R22/PhoSin_Vaquita2Epoch_1R22_Gamma_R22_Phocoena_sinus.mPhoSin1.pri.110_exons_DFE_plot.pdf}
        \caption{
        \label{fig:vaquita-dfe}
        Inferred versus simulated DFEs from simulated vaquita genomes.
        Simulations were performed using a two-epoch model of vaquita porpoise demography with a gamma-distributed DFE
        acting on nonsynonymous mutations with a relationship between the selection coefficient ($s$) and dominance coefficient ($h$) (see \textbf{Methods}).
        The DFE is inferred by analyzing all simulated genomes jointly by one of three different methods:
        \grapes, \polydfe, and \dadicli.
        (A-B) Mean absolute value of selection coefficient ($\lvert E(s) \rvert $) and shape parameter are
        shown for each DFE inferred from all three simulated datasets,
        with median values marked by horizontal bars
        and simulated values represented by dashed horizontal lines.
        (C) Binned distribution of $2hs$ implied by the simulated DFE compared with
        the distribution of $s$ implied by the average DFE inferred for each method from the three simulated datasets (averaging the inferred gamma parameters).
        The distribution of $2 h s$ is multimodal because of the simulated relationship between $h$ and $s$ (see text).
        }
    \end{figure}

@@

    The exploration of performance among methods done here is not meant to be thorough
    benchmarking, but rather to highlight the utility of the \stdpopsim resource.
    Having said that, there are a number of interesting results that merit discussion.
    The first pertains to estimation of demographic history ($N_e(t)$) in the context of
    linked purifying selection. Our simulations of human and vaquita genomes
    model deleterious mutations only at exonic positions of the genome.
    As shown in Figures \ref{fig:homsap-summary-stats} and \ref{fig:vaquita-summary-stats},
    this combination only produces sizable decreases in polymorphism at exonic positions,
    which represent a small minority of sites within the genome, and does not lead to large-scale
    perturbations in genome-wide diversity. Thus, it is not surprising that estimates of $N_e(t)$
    from genome-wide data from which exons have been masked should be accurate
    (e.g., Figures \ref{fig:human-demography} and \ref{fig:vaquita-demography}).
    Indeed our results echo those of \citet{marsh2024biases} who found \revpoint{AE}{1}
    similar results on realistic human simulations that included selection.
    However, the question remains to what extent this would be true in empirical data where
    selection is not constrained to coding positions. For instance, we know
    that many genomes harbor large numbers
    of conserved non-coding elements \citep[e.g.,][]{siepel2005evolutionarily} with extremely
    strong selection acting on them \citep{katzman2007human,mclean2008dispensability}.
    We thus caution regarding extrapolation from our results. Future studies might build upon
    the annotations we used here to include selection at other portions of the genome as well
    as mixtures of DFEs across sites.

    A second important finding is that inference of the DFE
    is sensitive to the dominance of deleterious mutations.
    All three of the DFE-inference methods applied here assume that deleterious mutations act
    in an additive fashion ($h=0.5$) by default.
    However, the simulated DFE for the vaquita genome included highly recessive mutations.
    Specifically, these simulations used a DFE where mutations with $s<-0.01$ are quite recessive,
    with $h<=0.01$, and where mutations with $-0.01<s<-0.001$ have $h=0.1$.
    We found that the three DFE-inference methods underestimated $E[s]$ and overestimated the shape parameter (Figure \ref{fig:vaquita-dfe}A-B).
    This performance is not surprising since decreasing $h$ for a given distribution of $s$ will lead to a decrease in $hs$,
    or an apparent decrease in the amount of selection.
    Thus, fitting a mis-specified DFE that assumes $h=0.5$ leads to underestimates of $E[s]$. \revpoint{1}{7}
    %IG: added this to respond to comment 1.7.
    %IG: I quote this in the response, so if we edit this, we should copy modified version to the response.
    Our analysis suggests that the distributions of $s$ estimated by these methods
    provide a reasonable approximation for the true distribution of $hs$ (Figure \ref{fig:vaquita-dfe}C).
    This points to a notable limitation of most existing DFE inference methods,
    which assume that selection coefficients are gamma-distributed.
    As shown in Figure \ref{fig:vaquita-dfe}C,
    while a gamma distribution might fit the distribution of $s$, a relationship between $h$ and $s$ might result in
    $hs$ which is not well approximated by a gamma distribution.

@@

    %IG: added this section to address rev's comment about lacking details in implementation
    \subsection*{Implementing a DFE model using \stdpopsim's \texttt{DFE} class}

@@

    Crucially, the advantageous allele must remain extant over the course of
    the sweep; this conditioning is accomplished via rejection sampling,
    whereby the simulation is restarted from the onset of the sweep if the
    allele drifts out of the population of origin (or alternatively, below a
    user-specified frequency). If there are competing advantageous alleles
    (simultaneous sweeps at distinct loci) and one of these drifts out of
    bounds, then the simulation is restarted from the onset of the first sweep;
    this avoids conditioning on particular (partial) allele frequency
    trajectories for the remaining alleles.  A downside of this strategy is
    that rejection becomes more likely with increasing numbers of simultaneous
    sweeps or with hard-to-satisfy constraints; but it allows for greater
    flexibility in sweep specification and for interoperability with the
    \stdpopsim DFE model.

@@

% constant size DFE figure
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/HomSap/Constant/HomSap_Constant_Gamma_K17_ensembl_havana_104_exons_DFE_plot}
    \caption{
       Inferred versus simulated DFEs from human genomes simulated with
       a simple demographic model with one population with a constant size of $10,000$
       and a gamma-distributed DFE acting on exons (see \textbf{Methods}).
        DFE is inferred by three different methods: \grapes, \polydfe , and \dadicli.
        Mean absolute value of selection coefficient ($\lvert E(s) \rvert$, panel A) and shape parameter (panel B)
        are shown for each DFE inferred from all three simulated datasets
        with median values marked by horizontal bars
        and simulated values represented by dashed horizontal lines.
    }
    \label{fig:homsap-dfe.constant}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/PhoSin/Constant/PhoSin_Constant_Gamma_R22_Phocoena_sinus.mPhoSin1.pri.110_exons_DFE_plot.pdf}
    \caption{
    \label{fig:vaquita-dfe.constant}
        Inferred versus simulated DFEs from vaquita genomes simulated with
        a simple demographic model with one population with a constant size of $3,500$
        and a gamma-distributed DFE acting on exons
        with a relationship between the selection coefficient ($s$) and dominance coefficient ($h$) (see \textbf{Methods}).
        DFE is inferred by three different methods: \grapes, \polydfe , and \dadicli.
        Mean absolute value of selection coefficient ($\lvert E(s) \rvert$, panel A) and shape parameter (panel B)
        are shown for each DFE inferred from all three simulated datasets
        with median values marked by horizontal bars
        and simulated values represented by dashed horizontal lines.
    }
\end{figure}

@@

>>>>

    % sweep interface
    While the \texttt{DFE} class facilitates the simulation of non-neutral variation across a sequence,
    it does not provide a means of specifying the origin and fate of particular selected alleles.
    Thus, we have also implemented an API in \stdpopsim (\texttt{stdpopsim.selective\_sweep}) that enables 
    selective sweeps to be simulated with a high degree of control.
    This function augments a demographic model by introducing an advantageous mutation over a given time interval,
    at a given position along the sequence; and conditions the simulation such that the mutation is above a particular 
    frequency at the beginning and end of the sweep period.
    The conditioning is implemented via rejection sampling; if the frequency of the advantageous allele 
    drifts out of the prescribed bounds, the simulation is restarted from the introduction of
    % DRS: not editing this because I haven't looked at the implementation, but this doesn't seem right.
    % Obviously the allele must be allowed to drift outside of the bounds during the simulation, as long as it finishes
    % within those bounds. So if it is fixed during the sim when the bounds don't extend to 1.0 (or lost when the
    % bounds don't extent to 0.0) then we restart, and we restart if at the end of the sweep period we are outside of
    % the bounds. right? If what I am saying is correct then maybe this should be rephrased, and more importantly the
    % corresponding part of the methods should be corrected.
    the advantageous allele. If there are competing advantageous alleles (e.g. simultaneous sweeps at distinct loci) 
    then the simulation is restarted from the introduction of the first of these.
    As this functionality is implemented on top of the existing \stdpopsim API, it may be combined with other 
    models of selection and demography to simulate the combined effects of multiple, disparate evolutionary processes
    on genetic variation in a population.

@@

    \begin{figure}[b!]
        \centering
        \includegraphics[width=\linewidth]{figures/HomSap/OOA/HomSap_discrete_DFE}
        \caption{Inferred versus simulated DFEs from simulated human genomes.
        Simulations were performed using a human out-of-Africa demographic model with a gamma-distributed DFE
        acting on exons (see \textbf{Methods}).
        DFE is inferred separately for each of the three extant populations (CEU, CHB, and YRI)
        by three different methods: \grapes, \polydfe , and \dadicli.
        (A-B) Mean absolute value of selection coefficient ($\lvert E(s) \rvert $) and shape parameter are
        shown for each DFE inferred from all three simulation replicates,
        with median values marked by horizontal bars
        and simulated values represented by dashed horizontal lines.
        (C) Binned distribution of $s$ implied by the average DFE inferred from the three simulated datasets (averaging the inferred gamma parameters);
        white bars represent the distribution used in the simulation. Standard error of the mean for each bin is shown in black. 
        }
        \label{fig:homsap-dfe.ooa}
    \end{figure}

@@

    \begin{figure}[b!]
        \centering
        \includegraphics[width=0.8\textwidth]{figures/PhoSin/Vaquita2Epoch_1R22/PhoSin_Vaquita2Epoch_1R22_Gamma_R22_Phocoena_sinus.mPhoSin1.pri.110_exons_DFE_plot.pdf}
        \caption{
        \label{fig:vaquita-dfe}
        Inferred versus simulated DFEs from simulated vaquita genomes.
        Simulations were performed using a two-epoch model of vaquita porpoise demography with a gamma-distributed DFE
        acting on nonsynonymous mutations with a relationship between the selection coefficient ($s$) and dominance coefficient ($h$) (see \textbf{Methods}).
        The DFE is inferred by analyzing all simulated genomes jointly by one of three different methods:
        \grapes, \polydfe, and \dadicli.
        (A-B) Mean absolute value of selection coefficient ($\lvert E(s) \rvert $) and shape parameter are
        shown for each DFE inferred from all three simulation replicates,
        with median values marked by horizontal bars
        and simulated values represented by dashed horizontal lines.
        (C) Binned distribution of $2hs$ implied by the simulated DFE compared with
        the distribution of $s$ implied by the average DFE inferred for each method from the three simulated replicates (averaging the inferred gamma parameters).
        The distribution of $2 h s$ is multimodal because of the simulated relationship between $h$ and $s$ (see text).
        }
    \end{figure}

@@

    The exploration of performance among methods done here is not meant to be thorough
    benchmarking. Rather, we include these analyses as examples to highlight the utility of \stdpopsim
    for both benchmarking and other tasks where researchers would benefit from the ability to perform
    simulations using parts of published population genetic models. in a streamlined and quality-controlled fashion.
    Having said that, there are a number of interesting results that merit discussion.
    The first pertains to estimation of demographic history ($N_e(t)$) in the context of
    linked purifying selection. Our simulations of human and vaquita genomes
    model deleterious mutations only at exonic positions of the genome.
    As shown in Figures \ref{fig:homsap-summary-stats} and \ref{fig:vaquita-summary-stats},
    this combination only produces sizable decreases in polymorphism at exonic positions,
    which represent a small minority of sites within the genome, and does not lead to large-scale
    perturbations in genome-wide diversity. Thus, it is not surprising that estimates of $N_e(t)$
    from genome-wide data from which exons have been masked should be accurate
    (e.g., Figures \ref{fig:human-demography} and \ref{fig:vaquita-demography}).
    Indeed our results echo those of \citet{marsh2024biases} who found \revpoint{AE}{1}
    similar results on realistic human simulations that included selection.
    However, the question remains to what extent this would be true in empirical data where
    selection is not constrained to coding positions. For instance, we know
    that many genomes harbor large numbers
    of conserved non-coding elements \citep[e.g.,][]{siepel2005evolutionarily} with extremely
    strong selection acting on them \citep{katzman2007human,mclean2008dispensability}.
    We thus caution regarding extrapolation from our results. Future studies might build upon
    the annotations we used here to include selection at other portions of the genome as well
    as mixtures of DFEs across sites.

    A second important finding is that inference of the DFE
    is sensitive to the dominance of deleterious mutations.
    All three of the DFE-inference methods applied here assume that deleterious mutations act
    in an additive fashion ($h=0.5$) by default.
    However, the simulated DFE for the vaquita genome included highly recessive mutations.
    Specifically, these simulations used a DFE where mutations with $s<-0.01$ are quite recessive,
    with $h<=0.01$, and where mutations with $-0.01<s<-0.001$ have $h=0.1$.
    We found that the three DFE-inference methods underestimated $E[s]$ and overestimated the shape parameter (Figure \ref{fig:vaquita-dfe}A-B).
    This performance is not surprising since decreasing $h$ for a given distribution of $s$ will lead to a decrease in $hs$,
    or an apparent decrease in the amount of selection.
    Thus, fitting a mis-specified DFE that assumes $h=0.5$ leads to underestimates of $E[s]$. \revpoint{1}{7}
    %IG: added this to respond to comment 1.7.
    %IG: I quote this in the response, so if we edit this, we should copy modified version to the response.
    Our analysis suggests that the distributions of $s$ estimated by these methods
    provide a reasonable approximation for the true distribution of $hs$ (Figure \ref{fig:vaquita-dfe}C).
    This points to a notable limitation of most existing DFE inference methods,
    which assume that selection coefficients are gamma-distributed.
    %DRS: Is the limitatio here that selection coefficients are gamma-distributed, or rather that h is fixed?
    As shown in Figure \ref{fig:vaquita-dfe}C,
    while a gamma distribution might fit the conditional distribution of $s$ given $h$, a relationship between $h$ and $s$ might result in a magrinal distribution for $hs$ which is not well approximated by a gamma distribution.

@@

    %IG: added this section to address rev's comment about lacking details in implementation
    %DRS: I think this section looks great.
    \subsection*{Implementing a DFE model using \stdpopsim's \texttt{DFE} class}

@@

    Crucially, the advantageous allele must remain extant over the course of
    the sweep; this conditioning is accomplished via rejection sampling,
    whereby the simulation is restarted from the onset of the sweep if the
    allele drifts out of the population of origin (or alternatively, below a
    user-specified frequency). If there are competing advantageous alleles
    % DRS: see my earlier comment about how we implement the conditioning here
    % and correct as needed
    (simultaneous sweeps at distinct loci) and one of these drifts out of
    bounds, then the simulation is restarted from the onset of the first sweep;
    this avoids conditioning on particular (partial) allele frequency
    trajectories for the remaining alleles.  A downside of this strategy is
    that rejection becomes more likely with increasing numbers of simultaneous
    sweeps or with hard-to-satisfy constraints; but it allows for greater
    flexibility in sweep specification and for interoperability with the
    \stdpopsim DFE model.

@@

% constant size DFE figure
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/HomSap/Constant/HomSap_Constant_Gamma_K17_ensembl_havana_104_exons_DFE_plot}
    \caption{
       Inferred versus simulated DFEs from human genomes simulated with
       a simple demographic model with one population with a constant size of $10,000$
       and a gamma-distributed DFE acting on exons (see \textbf{Methods}).
        DFE is inferred by three different methods: \grapes, \polydfe , and \dadicli.
        Mean absolute value of selection coefficient ($\lvert E(s) \rvert$, panel A) and shape parameter (panel B)
        are shown for each DFE inferred from all three simulation replicates
        with median values marked by horizontal bars
        and simulated values represented by dashed horizontal lines.
    }
    \label{fig:homsap-dfe.constant}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/PhoSin/Constant/PhoSin_Constant_Gamma_R22_Phocoena_sinus.mPhoSin1.pri.110_exons_DFE_plot.pdf}
    \caption{
    \label{fig:vaquita-dfe.constant}
        Inferred versus simulated DFEs from vaquita genomes simulated with
        a simple demographic model with one population with a constant size of $3,500$
        and a gamma-distributed DFE acting on exons
        with a relationship between the selection coefficient ($s$) and dominance coefficient ($h$) (see \textbf{Methods}).
        DFE is inferred by three different methods: \grapes, \polydfe , and \dadicli.
        Mean absolute value of selection coefficient ($\lvert E(s) \rvert$, panel A) and shape parameter (panel B)
        are shown for each DFE inferred from all three simulation replicates
        with median values marked by horizontal bars
        and simulated values represented by dashed horizontal lines.
    }
\end{figure}

@@

%sweep power vs exon density
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8 \textwidth]{figures/HomSap/sims_runtime_plot.pdf}
    \caption{
    Runtime of simulated human chromosomes with and without background selection and demography.
    Each point represents a single simulation of a single chromosome under conditions labeled by facets.
    }
    \label{fig:sim-runtime}
\end{figure}
